Inequality in socioemotional skills: a cross-cohort comparison

### 2019-02-28 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
new version of text and tables - incorporates most reviewer comments

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; less evidence there is however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}. Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{Blundell2007}.

@@

In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes. We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education. We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}). We present results with and without controlling for cognitive skills. As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts. Still, we control for a factor score that summarises all information on cognitive skills that is available in each cohort, regardless of their comparability.

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, he is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.

@@

For the British Cohort Study, we last examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.

>>>>

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; there is less evidence, however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006. It is likely that unmarried mothers in the two cohorts have very different characteristics. The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, as has been done for the analysis of wage inequality \citep{Blundell2007}.

@@

In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes. We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education. We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In \autoref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used. 'Raw' score, obtained by simple summing of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills. As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts. Still, we control for a factor score that summarises all information on cognitive skills that is available in each cohort, regardless of their comparability.

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, she is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.

@@

Cohort members from the British Cohort Study are now well into their adulthood. For this cohort, we can examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in  \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development. Similarly, \citet{Goodman2015} use data from the same cohort, but exploit a richer set of measures from the age 10 survey. They show that social and emotional skills measured at age 10 are strongly related to a wide range of adult outcomes.

### 2019-02-28 codes/R/paper/text/paper_jpube_4appendix.tex
new version of text and tables - incorporates most reviewer comments

In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}. The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'. Since they are all behaviours indicating lower skills, we encode all of them in reverse, i.e. `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2. We augment the 19-item Rutter Scale with four additional parent-reported questions from the parental questionnaire, items A to D. These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'. we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise. To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness"). We assign the lowest category among the two original items to the newly obtained item. We also recode items 5 and 14 to binary instead of three categories. These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.

@@

\section{Measurement invariance details}\label{asec:midet}

\subsection{Alternative parameterisations for the configural model}

There are infinite ways to parameterise the configural model defined by \eqref{eq:propens} and \eqref{eq:thresh}. Widely used parameterisations are:
\bi[label=$\diamond$]
\item \underline{Delta parameterisation} {[WE{\textDelta}]} \citep{Wu2016a}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$

\item \underline{Theta parameterisation} {[WE{\textTheta}]} \citep{Wu2016a}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$

\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\ei
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$). The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.

\subsection{Identification of models with different levels of invariance}

In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).

Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood. In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$). This makes identification reliant on more stringent normalisations. A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}. New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed. In light of this, they propose models that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts. A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.

>>>>

In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}. The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'. Since they are all behaviours indicating lower skills, we encode all of them in reverse, i.e. `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2. We augment the 19-item Rutter Scale with three additional parent-reported questions from the parental questionnaire, items A, B, and D. These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'. we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise. To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness"). We assign the lowest category among the two original items to the newly obtained item. We also recode items 5 and 14 to binary instead of three categories. These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.

@@

\section{Robustness of exploratory analysis}\label{asec:fullefa}

In this section, we repeat the exploratory analysis step in \autoref{sec:ea} for the full set of Rutter and SDQ items. This is to show that the factor structure emerging from exploratory analysis of the 11-item subscale is consistent with what would emerge considering the original scales in their entirety. Again, we proceed by first assessing the optimal number of factors, and then examining the loading obtained from exploratory factor analysis.

Results for the optimal number of factors as indicated by different approaches are in \autoref{tab:numfac_all}. Similarly to the 11-item subscale, there is not much agreement between methods. Since the purpose of this section is to assess the robustness of the 11-item subscale, we adopt a conservative approach by estimating EFA models with the largest number of factors suggested, i.e. five. In this way, we allow for richer factor solutions, that have more power to disprove our simpler two-factor solution for the novel subscale.

\autoref{tab:load_bcsall} presents factor loadings for the Rutter scale in BCS, with the addition of the "headaches/stomachaches" and "tantrums" items (see Appendix \autoref{asec:scales}). The split between externalising and internalising items that we recover in the 11-item scale is almost entirely preserved in the full scale, as seen by items loading on factors 1 and 2. The only exception is the "headaches/stomachaches" item, which seems to load on a separate factor. We carry out robustness checks for the measurement invariance analysis excluding this item in Appendix \autoref{asec:no511}.

The same analysis is repeated for the SDQ scale in MCS in \autoref{tab:load_mcsall}. An internalising, emotional dimension (factor 3) emerges neatly, and coherently with the analysis on our subscale. The externalising items from our subscales are split across two dimensions in this full-scale EFA: one more related to hyperactivity (factor 2) and one to conduct problems (factor 4). This is consistent with the original structure of the SDQ \citep{Goodman1997}.

\section{Robustness of item choice}\label{asec:no511}

In deriving our novel 11-item scale, we construct two items for the BCS cohort based on questions that are not in the original Rutter scale -- namely those concerning "headaches/stomachaches" and "tantrums" (see Appendix \autoref{asec:scales} above for details). Concerns might arise that introducing these items might somewhat invalidate our main conclusions, rather than provide additional informational content on children's externalising and internalising behaviours and symptoms.

In fact, exploratory factor analysis (on both on the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings. As a robustness check, we replicate the main results of the paper by excluding them from the subscale.

Panel C of \autoref{tab:fit} shows that the measurement invariance analysis yields the same qualitative results once these two items are included. \autoref{fig:scatter_no511} shows a scatter plot of the factor scores obtained from the factor model with and without items 5 and 11. They exhibit very high correlation, thus indicating that our results in \autoref{sec:results} would not substantially change if we omitted the two items with the least informational content.

### 2019-02-28 codes/R/paper/text/paper_jpube_5apptables.tex
new version of text and tables - incorporates most reviewer comments

\begin{table}[ht!]
\caption{Behavioural screening scales in the BCS and MCS five-year surveys}\label{tab:scales}
\centering\setstretch{1.2} \footnotesize
\begin{tabular}{L{.46\linewidth}L{.001\linewidth}L{.46\linewidth}}
\toprule
\multicolumn{3}{l}{\textbf{Panel A: Rutter A Scale \citep{Rutter1970}} -- British Cohort Study (1975) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.$^\dagger$
\item Is squirmy or fidgety.$^\dagger$
\item Often destroys own or others' belongings.
\item Frequently fights other children.$^\dagger$
\item Not much liked by other children.
\item Often worried, worries about many things.$^\dagger$
\item Tends to do things on his/her own, is rather solitary.$^\dagger$
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.$^\dagger$
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{12}
\item Frequently bites nails or fingers.
\item Is often disobedient.$^\dagger$
\item Cannot settle to anything for more than a few moments.$^\dagger$
\item Tends to be fearful or afraid of new things or new situations.$^\dagger$
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.$^\dagger$
\item[A.] Complains of headaches.$^\dagger$
\item[B.] Complains of stomach-ache or has vomited.$^\dagger$
\item[C.] Complains of biliousness.
\item[D.] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).$^\dagger$
\end{enumerate}
\\
\midrule
\multicolumn{3}{l}{\textbf{Panel B: Strength and Difficulties Questionnaire \citep{Goodman1997}} -- Millennium Cohort Study (2000/1) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Considerate of other people's feelings.
\item Restless, overactive, cannot stay still for long.$^\dagger$
\item Often complains of head- aches, stomach-ache or sickness.$^\dagger$
\item Shares readily with other children (treats, toys, pencils, etc.).$^+$
\item Often has temper tantrums or hot tempers.$^\dagger$
\item Rather solitary, tends to play alone.$^\dagger$
\item Generally obedient, usually does what adults request.$^{\dagger +}$
\item Many worries, often seems worried.$^\dagger$
\item Helpful if someone is hurt, upset or feeling ill.$^+$
\item Constantly fidgeting or squirming.$^\dagger$
\item Has at least one good friend.$^+$
\item Often fights with other children or bullies them.$^\dagger$
\item Often unhappy, down-hearted or tearful.$^\dagger$
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{13}
\item Generally liked by other children.$^+$
\item Easily distracted, concentration wanders.$^\dagger$
\item Nervous or clingy in new situations, easily loses confidence.$^\dagger$
\item Kind to younger children.$^+$
\item Often lies or cheats.
\item Picked on or bullied by other children.
\item Often volunteers to help others (parents, teachers, other children).$^+$
\item Thinks things out before acting.$^+$
\item Steals from home, school or elsewhere.
\item Gets on better with adults than with other children.
\item Many fears, easily scared.
\item Sees tasks through to the end, good attention span.$^+$
\end{enumerate}
\\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Items denoted by $^+$ are positively coded in the original scale. Items denoted by $^\dagger$ are retained in the new comparable scale.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
\begin{table}[ht!]
\caption{Description of harmonised variables}\label{tab:harmvar}
\centering\setstretch{1.2} \footnotesize
\begin{tabular}{L{.2\textwidth}C{.1\textwidth}L{.3\textwidth}L{.7\textwidth}}
\toprule
Variable Group & Age & Variable & Note \\
\midrule
Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth. School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972. \\ \\[-1.5em]
Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations. White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual). Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces. No father figure is a dummy for children whose father does not live in the same household. \\ \\[-1.5em]
Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old). Unmarried is only based on marital status, and includes cohabitation. \\ \\[-1.5em]
Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother. Gestational length and birth weight are from hospital records. \\ \\[-1.5em]
Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five. Three tests are used for BCS children: Copy Designs child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures). Three \emph{different} tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).

>>>>

\begin{table}[ht!]
\caption{Behavioural screening scales in the BCS and MCS five-year surveys}\label{tab:scales}
\centering\setstretch{1.2} \footnotesize
\begin{tabular}{L{.46\linewidth}L{.001\linewidth}L{.46\linewidth}}
\toprule
\multicolumn{3}{l}{\textbf{Panel A: Rutter A Scale \citep{Rutter1970}} -- British Cohort Study (1975) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.$^*$
\item Is squirmy or fidgety.$^*$
\item Often destroys own or others' belongings.
\item Frequently fights other children.$^*$
\item Not much liked by other children.
\item Often worried, worries about many things.$^*$
\item Tends to do things on his/her own, is rather solitary.$^*$
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.$^*$
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{12}
\item Frequently bites nails or fingers.
\item Is often disobedient.$^*$
\item Cannot settle to anything for more than a few moments.$^*$
\item Tends to be fearful or afraid of new things or new situations.$^*$
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.$^*$
\item[A.] Complains of headaches.$^*$
\item[B.] Complains of stomach-ache or has vomited.$^*$
\item[D.] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).$^*$
\end{enumerate}
\\
\midrule
\multicolumn{3}{l}{\textbf{Panel B: Strength and Difficulties Questionnaire \citep{Goodman1997}} -- Millennium Cohort Study (2000/1) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Considerate of other people's feelings.
\item Restless, overactive, cannot stay still for long.$^*$
\item Often complains of head- aches, stomach-ache or sickness.$^*$
\item Shares readily with other children (treats, toys, pencils, etc.).$^+$
\item Often has temper tantrums or hot tempers.$^*$
\item Rather solitary, tends to play alone.$^*$
\item Generally obedient, usually does what adults request.$^{* +}$
\item Many worries, often seems worried.$^*$
\item Helpful if someone is hurt, upset or feeling ill.$^+$
\item Constantly fidgeting or squirming.$^*$
\item Has at least one good friend.$^+$
\item Often fights with other children or bullies them.$^*$
\item Often unhappy, down-hearted or tearful.$^*$
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{13}
\item Generally liked by other children.$^+$
\item Easily distracted, concentration wanders.$^*$
\item Nervous or clingy in new situations, easily loses confidence.$^*$
\item Kind to younger children.$^+$
\item Often lies or cheats.
\item Picked on or bullied by other children.
\item Often volunteers to help others (parents, teachers, other children).$^+$
\item Thinks things out before acting.$^+$
\item Steals from home, school or elsewhere.
\item Gets on better with adults than with other children.
\item Many fears, easily scared.
\item Sees tasks through to the end, good attention span.$^+$
\end{enumerate}
\\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Items denoted by $^+$ are positively worded in the original scale. Items denoted by $^*$ are retained in the new 11-item comparable scale.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
\begin{table}[ht!]
\caption{Description of harmonised variables}\label{tab:harmvar}
\centering\setstretch{1.2} \scriptsize
\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
\toprule
Variable Group & Age & Variable & Note \\
\midrule
Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth. School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972. \\ \\[-1.5em]
Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations. White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual). Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces. No father figure is a dummy for children whose father does not live in the same household. Father's social class was recorded using the SOC2000 classification in MCS. We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class. \\ \\[-1.5em]
Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old). Unmarried is only based on marital status, and includes cohabitation. \\ \\[-1.5em]
Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother. Gestational length and birth weight are from hospital records. \\ \\[-1.5em]
Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five. Three tests are used for BCS children: Copy Designs child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures). Three \emph{different} tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).
